Summary of the things that I learned

### 2023-09-11 transcript/summary.tex
Write a summary of my work in week 36

>>>>

  \chapter{Week 36}
  \section{The Karoubi envelope}
  Given a category $ \CC $, define a category $ \overline \CC $, with objects being pairs $ (A, e) $, with $ e: A \to A $ an idempotent. A map $ (A, e), (B, f) $ is a map $ v: A \to B $ such that $ f \circ v \circ e = v $. The identity on $ (A, e) $ is $ e $, since if $ f \circ v \circ e = v $, then $ v \circ e = f \circ v \circ e \circ e = f \circ v \circ e = v $ (and likewise for $ f \circ v = v $).

  We can embed $ \CC $ into $ \overline \CC $ by sending $ A: \Ob \CC $ to $ (A, \mathop{Id}_A) $, and $ f: \Hom{\CC}{A}{B} $ to $ f $, since $ \mathop{Id}_B \circ f \circ \mathop{Id}_A = f $ is trivial.

  Given an idempotent $ e: \Hom{\CC}{A}{A} $, take $ r = e: \Hom{\overline \CC}{(A, \mathop{Id}_A)}{(A, e)} $ and $ s = e: \Hom{\overline \CC}{(A, e)}{(A, \mathop{Id}_A)} $. Then $ r \circ s = e \circ e = e = \mathop{Id}_{(A, e)} $ and $ s \circ r = e \circ e = e $. Therefore, $ e $ splits in $ \overline \CC $. This also shows that any object $ (A, e) : \Ob{\overline \CC} $ is a retract of (the image of) an object in $ \CC $. That is why the paper refers to $ \overline \CC $ as ``the category of retracts''.

  \section{Relatively cartesian closed}
  Hyland gives a new proof that $ \overline(\mathcal L(1)) $ is cartesian closed (when viewing $ \mathcal L(1) $ as a monoid with operation $ \bullet $, and then as a one-object category), based on the proof in \cite{taylor}.

  For a $ \Lambda $-algebra $ A $, we can take $ A(1) $, the set of functional elements of $ A $. This forms a monoid under composition: $ a; b = P a b $ for $ P = \lambda f g x, g(f x) $. The Karoubi completion of this category has as objects the terms $ a $ such that $ P a a = a $. The morphisms $ a \to b $ are elements $ f: A $ such that $ P a f = f = P f b $.

  Since we view $ \overline{A(1)} $ as a category, we can try to define products and exponentials of elements $ A\ B: \overline{A(1)} $. This is done in \S\S 1.3.5 and 1.3.6.

  Taylor defines a family $ (X_a)_{a: A} $ to be the \textit{display map} $ f: X \to A $ (the $ X_a $ arise as preimages/pullbacks of the $ a: A $). We have a category $ \CC^A $, consisting of families over $ A $. This is the slice category $ \CC / A $. If we have a morphism $ f: A \to B $, we get a functor $ \CC^f: \CC^B \to \CC^A $, which is the pullback along $ f $, and we will call it $ f^* $ or $ \texttt{P} f $. For a diagram
  \begin{center}
    \begin{tikzcd}
      Y \arrow[d] & X \arrow[d]\\
      A \arrow[r, "f"] & B
    \end{tikzcd}
  \end{center}
  there is a natural bijection between morphisms $ Y \xrightarrow{f} X $ over $ f $, and morphisms $ Y \to f^* X $.

  Pullbacks give binary products in $ \CC^A $. That is why these are also called ``fibred products''.

  Now, for indexed products: In $ \SET $, given indexed sets $ (X_a)_{a : A} $, we have the product $ \prod_{a: A} X_a $. Its elements are indexed families of elements $ (x_a)_{a: A} $. If we have an element $ X: \CC^A $, the elements of $ \prod_{a : A} X_a $ correspond to maps $ \Hom{\CC^a}{A}{X} $, i.e., maps that send $ a $ to an element in $ X_a $.

  The indexed product with a constant set $ \prod_{a: A} B $ is the exponential $ B^A $. Therefore, if a category has indexed products over itself, it is cartesian closed.

  We can also do this in an indexed fasion. If we have a family $ B $ indexed over $ A $ (given by a morphism $ f: A \to B $), and for every $ a: A $, a family $ (X_b)_{b : B_a} $ (equivalent to a morphism $ X \to B $), then we can construct the products $ (\prod \alpha X)_a := (\prod_{b: B_a} X_b) $ for all $ a : A $. That means that we have the type $ \prod \alpha X $ over $ A $.

  A morphism in $ \CC^A $ from $ C $ to $ (\prod \alpha X) $ is equivalent to a collection of morphisms $ \varphi_a: C_a \to (\prod \alpha X)_a = (\prod_{b: B_a} X_b) $.
  This is equivalent to a collection of morphisms $ \varphi_{ab}: C_a \to X_b $ for all $ a: A $ and $ b: B_a $.
  Note that $ a = \alpha b $, so $ C_a = C_{\alpha b} = (\alpha^* C)_b $.
  So for all $ b: B $, we need a morphism $ (\alpha^* C)_b \to X_b $. However, this is a morphism $ \alpha^* C \to X $ over $ B $.
  Therefore,
  \[ \Hom{\CC^A}{C}{\prod \alpha X} \cong \Hom{\CC^B}{\alpha^* C}{X} \]
  and the functor $ \prod \alpha: \CC^B \to \CC^A $ is a right adjoint to the functor $ \alpha^*: \CC^A \to \CC^B $.

  For a morphism $ B \to A $ and a morphism $ X \to B $, we can take the indexed sums $ (\sum \alpha X)_a := (\sum_{b: B_a} X_b)_a $.

  A morphism $ \sum \alpha X \to C $ in $ \CC^A $ is equivalent to a family of morphisms $ \varphi_a: \sum_{b: B_a} X_b \to C_a $.
  This is equivalent to, for every $ a: A $ and every $ b: B_a $, a morphism $ \varphi_{ab}: X_b \to C_a = (\alpha^* C)_b $.
  This is a morphism $ X \to \alpha^* C $ in $ \CC^B $, so $ \sum \alpha \dashv \alpha^* $.

  Therefore, Taylor defines an internal product in an indexed category to be a right adjoint to substitution over a display map, and an internal sum to be a left adjoint.

  \dots

  Hyland's view of this matter is as follows:

  Given a $ \lambda $-theory $ \mathcal L $, the category $ \mathbb R $ is the category of retracts of the idempotent elements of $ \mathcal L(0) $ (I think). That is: the objects of $ \mathbb R $ are elements $ a: \mathcal L(0) $, such that $ \lambda x, a (a x) = a $. Morphisms between idempotents $ a $ and $ b $ are elements $ f: \mathcal L(0) $ such that $ \lambda x, (b (f (a x))) = f $.

  \section{The Curry Festschrift}
  The Lambda calculus is a theory of functions. It specifies what one can do with functions (in a (concrete) cartesian closed category): One can abstract a function $ A \times B \to C $ to get a function $ A \to (B \to C) $; One can apply a function $ A \to B $ to an element of $ A $ to get an element of $ B $; And one always has the identity function $ x: A \to A $.

  For the typed lambda calculus, the objects in $ \CC $ form the types, and the morphisms (elements of the exponential object) are the lambda terms.

  For the untyped lambda calculus, we can say that all terms have the same domain and codomain $ U \to V $. However, since we can compose them with themselves, we must have $ U = V $. Also, given two terms $ f\ g: U \to U $, we can apply $ f $ to $ g $, so $ (U \to U) = U^U $ must be inside $ U $. Therefore, we want a section $ U^U \to U $ and a retraction $ U \to U^U $.

  \section{Comparing Hyland, Taylor and Curry}
  \subsection{Taylor}
  Taylor defines a \textit{Combinatory prealgebra} to be a set $ \Lambda $ with a binary operation $ \bullet $ (``application'') and elements $ K $ ($ \lambda x y, x $) and $ S $ ($ \lambda x y z, x z (y z) $) such that for all $ a\ b\ c: \Lambda $, $ K a b := K \bullet a \bullet b = a $ and $ S a b c = a c (b c) $.

  For $ f: \Lambda $, he defines $ f $ to be \textit{functional} if it has a representation $ f = \lambda x, a $.

  He defines a \textit{Combinatory algebra} to be a prealgebra $ \Lambda $ such that if $ fx = gx $ (in $ \Lambda [x] $, i.e., $ \Lambda $ to which we adjoin an additional element $ x $), then $ f = g $.

  He defines a \textit{model} to be a combinatory algebra $ \Lambda $, such that for all $ f\ g: \Lambda $, if for all $ a: \Lambda $, $ f a = g a $, then $ f = g $.

  He then defines $ \mathbf{Retr}(\Lambda) $ to be the category of retracts of the monoid of functional elements in $ \Lambda $. An object is an element $ A: \Lambda $ such that $ P A A = A $, and a morphism $ A \to B $ is an element $ \alpha: \Lambda $ such that $ P A \alpha = \alpha = P \alpha B $.

  % He notes that $ \mathbf{Retr}(\Lambda) $ is concrete iff $ \Lambda $ is a combinatory model (iff the monoid has enough constants).
  He notes that we can view objects (idempotents) of $ \mathbf{Retr}(\Lambda) $ as types $ \Vert A \Vert = \{ a \mid a = A a \} \subseteq \Lambda $ and morphisms $ \alpha: \mathbf{Retr}(\Lambda)(A, B) $ as maps $ a \mapsto \alpha a: \Vert A \Vert \to \Vert B \Vert $.
  % Take $ K = \lambda x y, x $ and $ \bot = (\lambda x, x x) (\lambda x, x x) $. Take $ T = K \bot = \lambda y, (\lambda x, x x) (\lambda x, x x) $. Then $ \Vert T \Vert = \{ \bot \} $ and every type $ A $ has a ``least'' element $ A \bot $.

  At the end of chapter $ 1 $, he shows that $ \mathbf{Retr}(\Lambda) $ is cartesian closed.

  In chapter 4, Taylor introduces a fibration. If $ \CC $ is a category and $ D $ is a class of morphisms, we say that $ D $ is a class of \textit{Display Maps} for $ \CC $ if $ D $ is closed under pullbacks along $ \CC $-morphisms, under composition and the maps to the terminal object are in $ D $.

  Then, given any $ A: \mathcal C $, we can construct $ \CC /_{\mathcal D} A $ as a full subcategory of $ \CC / A $, containing as objects only the display maps to $ A $. The $ \CC /_{\mathcal D} A $ form a fibration over $ \CC $.

  For $ \alpha: \CC(A, B) $, we have $ \alpha_*: \CC /_{\mathcal D} A \to \CC /_{\mathcal D} B $ ("the substitution functor"). It has a left adjoint, given by postcomposition, which we will call $ \sum \alpha $. If it also has a right adjoint, (which we then call $ \prod \alpha $), we call the category relatively cartesian closed.

  Taylor notes that $ \CC /_{\mathcal D} 1 $, the fiber over the terminal object, consists of all of $ \CC $. He also notes that a category that is relatively cartesian closed has cartesian closed fibers.

  \subsection{Hyland}
  Hyland defines an algebraic theory to be a functor $ \mathcal T: F \rightarrow \SET $, together with variables $ \pi_i : \mathcal T(n) $ for all $ 1 \leq i \leq n $, and an operation (``substitution'') $ \bullet: \mathcal T(m) \times \mathcal T(n)^m \to \mathcal T(n) $ which is associative, unital, dinatural.

  He defines a $ \lambda $-theory to be an algebraic theory $ \mathcal L $, together with retractions $ \rho: \mathcal T(n) \to \mathcal T(n + 1) $, that are compatible with $ \bullet $ (which, on $ \mathcal T(n + 1) $, leaves alone the $ n + 1 $th component).

  He defines $ \Lambda $ to be the $ \lambda $-theory corresponding to the $ \lambda $-calculus, with $ \Lambda(n) $ the set of lambda terms in $ n $ indeterminates.

  Finally, he defines a $ \Lambda $-algebra (one would call this a ``model'') to be a set $ A $, together with an action $ \mathcal \Lambda(n) \times A^n \to A $ for all $ n $ that is associative and unital.

  \subsection{Curry}

  Curry does not give a formal characterization of the $ \lambda $-calculus (he refers to `Lambek's paper'). However, he notes that there is a correspondence between cartesian closed categories, and `extensional typed $ \lambda $-calculi'.

  This means that a typed $ \lambda $-calculus has at least an identity function $ x_1: A \to A $. He also notes that, given a term $ f: A \times B \to C $, we can abstract it to $ \lambda b, f: A \to (B \to C) $. Finally, given variables $ u: A \to B $ and $ v: A $, we have $ u (v): (A \to B) \times A \to B $. Therefore, we have abstraction and application.

  He also notes that we need extensionality: if $ f = g $, then $ \lambda x, f = \lambda x, g $.

  For the untyped $ \lambda $-calculus, he constructs the terms `in the usual way', by starting with variables $ x_1, \dots $ and then using abstraction and application.

  He then constructs the category $ \mathbb R $ straightforwardly, with as objects the terms $ A $ without free variables, such that $ A = \lambda x, A (A x) $. The maps $ f: A \to B $ are terms $ f $ without free variables, for which we can prove $ f = \lambda x, B (f (A x)) $.

  He also constructs products $ A \times B := \lambda u z, z(A(u(\lambda x y, x)))(B(u(\lambda x y, y))) $, with projections
  \[ p_{AB} := \lambda u, (A \times B)(u)(\lambda xy, x) \quad \text{and} \quad q_{AB} := \lambda u, (A \times B) u (\lambda xy, y), \]
  with the product of morphisms $ f: C \to A $ and $ g: C \to B $ being $ \langle f, g \rangle := \lambda t z, z(f t)(g t) $.

  Finally, he defines the function space to be
  \[ A \to B := \lambda f, B \circ f \circ A \]
  with evaluation
  \[ \varepsilon_{BC} = \lambda u, C(u(\lambda x y, x))(B(u(\lambda x y, y))) \]
  and for $ h: A \times B \to C $,
  \[ \Lambda_{ABC} h = \lambda x y, h (\lambda z, z x y). \]
  He takes the reflexive object $ U := \lambda x, x $.
